---
title: "ProjectFirst"
author: "R Piazza"
date: "January 9, 2016"
output: html_document
---

Overall, we'll be looking at summary statistics in our activity dataset
First, we'll read in and process the data 

```{r, echo = TRUE}
library(ggplot2)
#First read it in
act <- read.csv(unzip('activity.zip'))

#Now the dates to date format  
act$date <- as.Date(as.character(act$date))
```

Now I'll make a histogram of the total number of steps taken each day 

```{r}
SumOfSteps <- aggregate(steps ~ date, act, sum)
qplot(steps, data = SumOfSteps)
```

Let's answer the question of what is the mean total number of steps taken per day?
```{r}
mean(SumOfSteps$steps)
```

How about the median?]
```{r}
median(SumOfSteps$steps)
```

What does a typical day look like? 
```{r}
#Time series plot of the average number of steps taken  
AvgActivityPattern <- aggregate(steps ~ interval, act, mean)  
qplot(interval, steps,data = AvgActivityPattern, geom = "line")  
```

What is the 5-minute interval that, on average, contains the maximum number of steps?
```{r}
AvgActivityPattern$interval[AvgActivityPattern$steps== max(AvgActivityPattern$steps)]  
```

Now, we have some missing data, to impute it, I'll take the overall average value for that 5 minute period and replace the NA with it.
```{r}
#first find where we have NA values  
mean(is.na(act$steps)) 
 
mean(is.na(act$date))   
 
mean(is.na(act$interval))  
#how many NA's are we talking about?  
sum(is.na(act$steps)) 
```

So, we have 2304 rows or 13% of the steps values are NA and steps is the only column we have to imput. 

I'll go line by line, taking the first NA value found, 
Lookup the interval associated, 
Lookup that associated average activity in AvgActivityPattern and 
replace the corresponding step value

```{r}
#First create a new dataset
ActImputed <- act
#just keep going through the dataset until no more NA are found
while (mean(is.na(ActImputed$steps) != 0)) {
    StepsNA <-is.na(ActImputed$steps)
    MatchesOverallAverage <- AvgActivityPattern$interval== ActImputed$interval[StepsNA][1]
    ActImputed$steps[StepsNA][1] <- AvgActivityPattern$steps[MatchesOverallAverage]
}
```

Make a new histogram of the total number of steps taken each day with imputed data
```{r}
SumOfStepsImputed <- aggregate(steps ~ date, ActImputed, sum)
qplot(steps, data = SumOfStepsImputed)
```

What's the new mean and median?
```{r}
mean(SumOfStepsImputed$steps) #10766.19 Same as before
median(SumOfStepsImputed$steps) #10766.18 Now went up slightly and is also the same as the mean
```
So the mean stayed the same but the median changed by a little and also is very close to the mean


Let's create a panel plot comparing the average number of steps taken per 5-minute interval across weekdays and weekends
```{r}
#add another column to Activity Imputed called Weekday and set is as weekday or weekend
ActImputed$weekday[weekdays(ActImputed$date) == 'Saturday' | weekdays(ActImputed$date) == 'Sunday'] <- 'weekend'
ActImputed$weekday[weekdays(ActImputed$date) != 'Saturday' & weekdays(ActImputed$date) != 'Sunday'] <- 'weekday'
#change all weekends and weekdays to factors
ActImputed$weekday <- as.factor(ActImputed$weekday)
#create the plot and color according to weekday
MeanWeekday <- aggregate(steps ~ interval+weekday, data = ActImputed, FUN = mean)
qplot(x = interval, y = steps, data = MeanWeekday, facets = weekday~., geom = "line", col = weekday)
```